### 2019-05-22 chapter_00_open_science.tex
Bullets in python

We shall now discuss these topics in greater detail. The chapter is organized
into two sections. The first section focusses on the technical aspects, with
particular focus on emerging developments in programming languages, tools and
services suited for open-source software development. We shall present Python's
strengths and weaknesses as a language for scientific computing.
%
The section ends with a note on the capabilities of FluidDyn
packages\footnote{We use FluidDyn (with capital letters) to name the project
and \fluiddyn for the base package.}.
%
In the second section a comparison of various software development
methodologies are made. Here, we will also discuss about the possible
contradiction between productivity of individuals and productivity at the
community level.
%
% The chapter finally ends with a summary of the motivations for the project
% FluidDyn and an overview of the packages used.

@@

\subsubsection{Communication}

@@

%\paragraphbf{Automatic web documentations built on servers}

% We need transitions and some words at the end of the subsections to help the
% readers to understand why we present what we present and what is the link
% with the FluidDyn project...

Thus, there is a number of new methods and tools in software development.
Taken all together, they greatly improve our collective efficiency and
open possibilities of interactions that were unthinkable a few years ago.
%
For example, the success of community-driven software like \pack{astropy} (a
subject-area research library with
\href{https://github.com/astropy/astropy}{more than 240 contributors}) and
\pack{scikit-learn} (a machine learning toolkit with
\href{https://github.com/scikit-learn/scikit-learn}{more than 1000
contributors}) would not have been possible without these new software
development methods.
%
% pa: done: we need to mention in this paragraph existing community-driven
% subject-area research libraries (see the file
% ./communitydriven_research_soft.md).
%
As scientists, we can wonder how to fully exploit these new possibilities to do
science.  The project FluidDyn is an attempt to help the fluid dynamics
community to advance in this direction.

% TODO:

@@

Python community has thrived remarkably in the last few years, evident from the
fact that by the end of 2017, Python became the second most popular language in
terms of pull-requests in GitHub\footnote{See GitHub's annual report titled
\href{https://octoverse.github.com/}{ ``The State of the Octoverse 2017''}.} and
questions on StackOverflow\footnote{See
\href{https://stackoverflow.blog/2017/09/06/incredible-growth-python/}{ ``The
Incredible Growth of Python''}, a blog post by Stack Overflow.}.
%
TIOBE-index\footnote{\url{https://www.tiobe.com/tiobe-index}}, which assess the
popularity of a language based on number of skilled users, courses and
third-party vendors, ranks Python at number four. IEEE Spectrum\footnote{See
the analysis titled \href{https://spectrum.ieee.org/computing/software/%
the-2017-top-programming-languages}{``The 2017 Top Programming Languages''} on
IEEE spectrum.} ranks Python as the most popular language in 2017 with a metric
based on trends, demand by employers and users on open-source hubs.  All the
rankings do agree that Python usage has been steadily growing and is here to
stay.  Python largely outclasses languages like Fortran and Matlab according to
these statistics, and is now clearly one of the mainstream languages with C,
Java, C++ and Javascript.
%
Let us summarize some characteristics of the Python languages that have lead to
such incredible success.

The Python language has been designed to boost the communication of technical
ideas between humans. The resulting code is as simple as possible due to the
inherent emphasis on readability. Humans can focus on the ideas. Python differs
from other mainstream languages for a variety of reasons:

\begin{itemize}
\item \textbf{Aesthetics.} The syntax is nice and elegant. Blocks of code are
defined with the indentation.
\item \textbf{Code style.} There are explicit guidelines
(\href{https://www.python.org/dev/peps/pep-0008/}{PEP~8}) supporting the
regularity (and thereby readability) of the code and tools
(linters) to evaluate adherence to the style.
\item \textbf{Dynamic typing.} The variables (the ``names'' in the code) are
not attached for life to an object in memory\footnote{Opposite behavior
compared to static languages as Fortran, C and C++.}. The types of the objects
are inferred at run time from the code and the context, and in many situations,
one does not need to declare them explicitly.
%
% Being a strongly typed language\footnote{In contrast to weakly typed languages,
% such as JavaScript}, types of the objects are respected and implicit type
% conversions are rarely performed.
% av: "rarely / never"?
% eg: 'hello' * 2 -> 'hellohello'; 1/3 -> 0.3333333
%
% pa: I think we can not cover such complicate notion in this document... It is
% anyway not so important for our demonstration.
\item \textbf{Automatic memory management.}  The user need not handle the
memory allocation and deallocation by hand.  There are, of course, ways to
control the memory usage of a program, though not as precisely as for low-level
languages.
\item \textbf{Interpreted.} The standard way\footnote{There are also tools to
compile Python code (for example Cython and Pythran).} to run a Python code is
to ``interpret'' it.  A program called the interpreter executes the code
nearly\footnote{Actually, the execution of Python code is less simple and
involves an Abstract Syntax Tree.}  instruction-by-instruction, with very few
optimizations.
%
In contrast to other compiled languages such as Fortran, C or C++, there is no
proper compilation step: the code is not translated to optimized machine
instructions, but only to Python bytecode.
%
We will discuss shortly the disadvantages of this characteristic. Two evident
advantages are rapid development cycle and the possibility of interactive
workflow (usually with IPython or Jupyter notebooks).

Note that there are many implementations of Python interpreters\footnote{We can
cite CPython (written in C), Jython (Java), IronPython (C\#), PyPy (Rpython, a
subset of Python) and MicroPython/CircuitPython (C, targeted to
micro-controllers).}. The default and most widely used implementation is written
in C and is called CPython.

\item \textbf{Simple, yet powerful.}

\begin{itemize}
\item Python supports multiple programming paradigms, including imperative,
object-oriented and functional.

\item It is easy to interface with code written in other languages (in particular
C, C++ and Fortran).

\item Python comes with a large and high quality \href{https://docs.python.org%
/3/library/index.html}{standard library}.

\item Python can be deployed on many different machines with different operating
systems (Linux, Windows, macOS, Android) and architectures (from a microcontroller
--- with MicroPython or CircuitPython --- to a Blue Gene supercomputer).

\end{itemize}
\end{itemize}

Thus, Python is one of the best ``glue'' languages for fast prototyping.
Developing with Python is much faster and easier than with many other
languages.  Less bugs are introduced by the developers simply because there are
much less lines of code \cite{nanz_comparative_2015}.
%
Maintaining and modifying existing code is also eased by the very good code
readability associated with the language.

Another consequence of the apparent simplicity of Python is that the learning
curve has a nice shape. Learning Python is very easy at the beginning and the
complexity gradually increases as the level of the developer increases.
%
Therefore, Python is good for developers of all levels: very gentle for
beginners and very powerful for advanced users.

Python is an old language (first implementation in 1991!) but it continues to
evolve.
\begin{itemize}
\item A hard and controversial decision was taken to clean up Python from its
historical incoherences by having a new series of backwards-incompatible
versions of the language, Python 3.  Python 3.0 was released in December 2008.
After a long and difficult process, the transition from Python 2 to Python 3 is
now (2018) behind us for nearly all important packages\footnote{Close to
\href{https://python3wos.appspot.com/}{95\% of the most downloaded packages}
are Python 3 compatible now.}.  Especially for scientific applications, we can
now work only in Python 3. For instance, \href{https://github.com/%
numpy/numpy/blob/master/doc/neps/dropping-python2.7-proposal.rst}{future
releases of \Numpy} and of \href{http://www.python3statement.org/}{the main
scientific packages} will soon drop compatibility with Python 2. These changes
open doors to very clean, coherent and potentially faster Python
experience. Therefore, one should not use Python 2 for science anymore and
instead, adopt the newer versions ($>=3.5$ in 2018).

\item New versions of Python have nice new features to adapt to new usages and to
recent trends in computer science.  For example, three new features were
introduced in Python 3.5 (first released on September 2015): the \codeinline{@}
operator for matrix multiplication, the new \codeinline{async} and
\codeinline{await} keywords for concurrency and type hinting\footnote{See also our
paragraph on the issue of type checking in the following.} (with the module
\codeinline{typing} and an associated syntax presented in the
\href{https://www.python.org/dev/peps/pep-0484/}{PEP~484}).
\end{itemize}

\paragraphbf{An incredible success}

Being a very versatile language, Python is widely used for many different
applications:
\begin{itemize}
\item Simple scripting.

\item System, database and network administration.

\item GNU/Linux distribution software.

\item Python is now one of the main scripting languages to add programmability
to applications (for example Paraview, Visit, QGIS, Blender, ...).

\item Web servers.

\item Web scraping and data analysis.

\item Animation movies, game development and gaming.

\item Education. Python has recently become one of the most popular language
for teaching programming and computer science.

\item Science!
Python is now widely used in scientific applications. It has now a mature and
powerful scientific ecosystem with well-established based packages (\Numpy for
N-dimensional homogeneous arrays, \Scipy as the fundamental toolkit for
scientific computing, \pack{Matplotlib} for plotting and \pack{Pandas} for data
structures) and several more specialized packages (to name a few, \pack{h5py},
\pack{mpi4py}, \pack{skimage}, \pack{sklearn}, ...). There are now great (and
often shiny) tools for most of the applications. For example, the
\href{https://github.com/rougier/python-visualization-landscape}{%
landscape of visualization tools usable through Python} is impressive.
%
Python is one of the main languages for data science with packages such as
\pack{Pandas}, \pack{statmodels}, \pack{sklearn}, \pack{Keras}, \pack{TensorFlow}
and \pack{PyTorch}.

There are now ready-to-use ``batteries-included'' Python installers
(similar to Matlab), like in particular the
\href{https://en.wikipedia.org/wiki/Anaconda_(Python_distribution)}{%
freemium open-source distribution Anaconda}, so it is very easy to start using
Python for scientific purposes.

There are now nice Python integrated development environments (IDE) adapted for
scientists, for example \href{https://github.com/spyder-ide/spyder}{Spyder},
\href{http://www.pyzo.org/}{Pyzo} and
\href{https://jupyterlab.readthedocs.io}{Jupyterlab}.

@@

>>>>

We shall now discuss these topics in greater detail. The chapter is organized
into three sections. The first section focusses on the technical aspects, with
particular focus on emerging developments in programming languages, tools and
services suited for open-source software development. In the next section, we
present Python's strengths and weaknesses as a language for scientific
computing.
%
The section ends with a note on the capabilities of FluidDyn
packages\footnote{We use FluidDyn (with capital letters) to name the project
and \fluiddyn for the base package.}.
%
The last section compares various software development methodologies. Here, we
will also discuss about the possible contradiction between productivity of
individuals and productivity at the community level.
%
% The chapter finally ends with a summary of the motivations for the project
% FluidDyn and an overview of the packages used.

@@

\subsubsection{Knowledge hubs and communication}

@@

Thus, there is a number of new methods and tools in software development.
Taken all together, they greatly improve our collective efficiency and
open possibilities of interactions that were unthinkable a few years ago.
%
For example, the success of community-driven software like \pack{astropy} (a
subject-area research library with
\href{https://github.com/astropy/astropy}{more than 240 contributors}) and
\pack{scikit-learn} (a machine learning toolkit with
\href{https://github.com/scikit-learn/scikit-learn}{more than 1000
contributors}) would not have been possible without these new software
development methods.
%
As scientists, we can wonder how to fully exploit these new possibilities to do
science.  The project FluidDyn is an attempt to help the fluid dynamics
community to advance in this direction.

@@

The Python community has thrived remarkably in the last few years, which is
evident from the fact that by the end of 2017, Python became the second most
popular language in terms of pull-requests in GitHub\footnote{See GitHub's
	annual report titled \href{https://octoverse.github.com/}{ ``The State
of the Octoverse 2017''}.} and questions on StackOverflow\footnote{See
\href{https://stackoverflow.blog/2017/09/06/incredible-growth-python/}{ ``The
Incredible Growth of Python''}, a blog post by Stack Overflow.}.
%
TIOBE-index\footnote{\url{https://www.tiobe.com/tiobe-index}}, which assess the
popularity of a language based on number of skilled users, courses and
third-party vendors, ranks Python at number four. IEEE Spectrum\footnote{See
	the analysis titled
	\href{https://spectrum.ieee.org/computing/software/%
the-2017-top-programming-languages}{``The 2017 Top Programming Languages''} on
IEEE spectrum.} ranks Python as the most popular language in 2017 with a metric
based on trends, demand by employers and users on open-source hubs.  All the
rankings agree that Python usage has been steadily growing and is here to stay.
Python largely outclasses languages like Fortran and Matlab according to these
statistics, and is now clearly one of the mainstream languages with C, Java,
C++ and JavaScript.
%
The language has been designed to boost the communication of technical ideas
between humans. 
%
Let us summarize some characteristics of the Python languages that have lead to
such an incredible success.
%
Python differs from other mainstream languages for a variety of reasons:

\begin{itemize}
\item \textbf{Aesthetics.} The syntax is expressive, with an inherent emphasis
on readability. Blocks of code are defined with the indentation.

\item \textbf{Code style.} There are official (but, optional) guidelines, named
\fnref{https://www.python.org/dev/peps/pep-0008/}{PEP~8}, advocating for
regularity of the code. Linters and code-formatters can evaluate adherence to
this style.

\item \textbf{Automatic memory management.}  A garbage collector handles the
memory allocation and deallocation, which needs no user intervention. There
are, of course, ways to control the memory usage of a program, though not as
precisely as for low-level languages.

\item \textbf{Dynamic typing.}
The types of the objects are inferred at run time from the code and the
context, and in many situations, one does not need to declare them explicitly.

\item \textbf{Interpreted.} The standard way\footnote{There are also tools to
compile Python code (for example Cython and Pythran).} to run a Python code is
to ``interpret'' it.  A program called the interpreter executes the code
nearly\footnote{Actually, the execution of Python code is less simple and
involves an Abstract Syntax Tree.} instruction-by-instruction, with very few
optimizations.
%
In contrast to other compiled languages such as Fortran, C or C++, there is no
proper compilation step: the code is not translated to optimized machine
instructions, but only to Python bytecode.
%
Two evident advantages are rapid development cycle and the possibility of
interactive workflow (usually with IPython or Jupyter notebooks).
%
Note that there are many implementations of Python
interpreters\footnote{Notable ones are CPython (written in C), Jython (Java),
IronPython (C\#), PyPy (Rpython, a subset of Python) and
MicroPython/CircuitPython (C, targeted to micro-controllers).}. The default and
most widely used implementation is written in C and is called CPython.

\item \textbf{Multi-paradigm.} Supports imperative, object-oriented, functional
and aspect-oriented programming.

\item \textbf{Extension support}. It is easy to interface with code written in
several other languages (in particular C, C++ and Fortran).

\item \textbf{Batteries included.} Python comes with a large and high quality
\href{https://docs.python.org% /3/library/index.html}{standard library}.

\item \textbf{Cross platform.} Python can be deployed on many different
machines with different operating systems (Linux, Windows, macOS, Android) and
architectures (from a microcontroller --- with MicroPython or CircuitPython ---
to a Blue Gene supercomputer).

@@

Thus, Python is one of the best ``glue'' languages for fast prototyping.
Code development in Python is much faster and easier than with many other
languages. Less bugs are introduced by the developers simply because there are
fewer lines of code \citep{nanz_comparative_2015}.
%
Maintaining and modifying existing code is also eased by the very good code
readability associated with the language.
%
Another consequence of the apparent simplicity of Python is that the learning
curve has a nice shape. Learning Python is very easy at the beginning and the
complexity gradually increases as the level of the developer increases.
%
Therefore, Python is good for developers of all levels: very gentle for
beginners and very powerful for advanced users.

Python is an old language (first implementation in 1991!) but it continues to
evolve.
%
A hard and controversial decision was taken to clean up Python from its
historical incoherences by having a new series of backwards-incompatible
versions of the language, Python 3.  Python 3.0 was released in December 2008.
After a long and difficult process, the transition from Python 2 to Python 3 is
now (2018) behind us for nearly all important packages\footnote{Close to
\href{https://python3wos.appspot.com/}{95\% of the most downloaded packages}
are Python 3 compatible now.}.  Especially for scientific applications, we can
now work only in Python 3. For instance, \href{https://github.com/%
numpy/numpy/blob/master/doc/neps/dropping-python2.7-proposal.rst}{future
releases of \Numpy} and of \href{http://www.python3statement.org/}{the main
scientific packages} will soon drop compatibility with Python 2. These changes
open doors to very clean, coherent and potentially faster Python
experience. Therefore, one should not use Python 2 for science any more and
instead, adopt the newer versions.

New versions of Python have nice new features to adapt to new usages and to
recent trends in computer science.  For example, three new features were
introduced in Python 3.5 (first released on September 2015): the \codeinline{@}
operator for matrix multiplication, the new \codeinline{async} and
\codeinline{await} keywords for concurrency and type hinting\footnote{See also our
paragraph on the issue of type checking in the following.} (with the module
\codeinline{typing} and an associated syntax presented in the
\href{https://www.python.org/dev/peps/pep-0484/}{PEP~484}).

Being a very versatile language, Python is widely used for many different
applications\footnote{\href{https://www.jetbrains.com/research/devecosystem-2018/python/}{See
\emph{The State of Developer Ecosystem Survey in 2018} by JetBrains}}.

\begin{itemize}
\item Simple scripting

\item Web development

\item Data science

\item System, database and network administration

\item GNU/Linux distribution software

\item Python is now one of the main scripting languages to add programmability
to applications (for example Paraview, Visit, QGIS, Blender, ...)

\item Web scraping

\item Animation, gaming and film industry

\item Education

\item Science

Python has recently become one of the most popular language
for teaching programming and computer science.
Python is now widely used in scientific applications. It has now a mature and
powerful scientific ecosystem with well-established based packages (\Numpy for
N-dimensional homogeneous arrays, \Scipy as the fundamental toolkit for
scientific computing, \pack{Matplotlib} for plotting and \pack{Pandas} for data
structures) and several more specialized packages (to name a few, \pack{h5py},
\pack{mpi4py}, \pack{skimage}, \pack{sklearn}).
The \href{https://github.com/rougier/python-visualization-landscape}{%
landscape of visualization tools usable through Python} makes it possible to
integrate post-processing and analysis capabilities into a single application.
%
Python is one of the main languages for data science with packages such as
\pack{Pandas}, \pack{statmodels}, \pack{sklearn}, \pack{Keras}, \pack{TensorFlow}
and \pack{PyTorch}.

There are now ready-to-use ``batteries-included'' Python installers (similar to
Matlab), like in particular the
\href{https://en.wikipedia.org/wiki/Anaconda_(Python_distribution)}{%
open-source distribution Anaconda}, so it is very easy to start using Python
for scientific purposes. Integrated development environments (IDE) adapted for
scientists, for example \href{https://github.com/spyder-ide/spyder}{Spyder},
and \href{https://jupyterlab.readthedocs.io}{JupyterLab}.
